---
title: "Latimer_phenology_notebook"
author: "Data collected by John Latimer<br/>data annotated by xxxxx <br/> code contributions by David J. Weston<br/>Travis Lawrence"
date: "12/01/2018"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
  pdf_document:
    toc: yes
---

# Introduction

The goal of this tutorial is to intorduce the public to the importance of community science, along with its inherent limitations. 

We are using a dataset collected by John Latimer for over 30 years. The data is publically available and the specific data used for this tutorial is available at GITHUB address......

# data description including caveates

```{r setup, include=FALSE}

library(captioner)

fig_nums <- captioner(prefix = "Fig.")
fig.1_cap <- fig_nums(name = "fig_1", 
                        caption = "2018 data, change in growth (initial weight (wt) - final wt among the two symbiotic partners alone and when added togather but grown seperately (_seperate) and when grown together in same well (_symbiosis). Cyano in symbiosis includes, endophytes, epiphytes and free living cyano.  Data incuded all pH conditions")

```


# 2018 Data import and initials glimpse


The origianl data provided by Latimer is a typical date format. However, ecolgoists and climate scientist often use Julian day format to provide a continuous count of the days starting at January 1st each year. Each year will contain 365 days or 366 days if a leap years

Code for changing the date format to the Julian day format is modified from the Neon site: https://www.neonscience.org/julian-day-conversion-r

this code nicely accounts for leap years. 


```{r, echo=FALSE, message=FALSE}
library(lubridate)
library(tidyverse)

datphen<-read.csv("Latimer_Cleaned_1.csv")
#str(datphen) #describes data class of columns within dataframe
#dim(datphen) #dimension of the dataframe

# From str(datphen) output see that 'date' is factor. change to 'date' class:

datphen$date<- as.Date(datphen$date, "%m/%d/%y")
#str(datphen) #to confirm class change


#Add columns for month, year and julian day to datphen
datphen2<-datphen %>% mutate(month = format(date, "%m"), year = format(date, "%Y"), julian = yday(date))

#quick data checkt to confirm new columns are OK especiallu given leap years.
datphen2 %>% group_by(year) %>% summarize(max(julian), min(julian))

```

We can see from the data that John did not record observations daily. However, the data provided span 1984 to 2016 for 32 years of observations!

As a quick check on the Julian day conversion, I looked at 06-27 julian days. on regular years it is 178, on leap years (2000,2004,2008,2012, etc) it is 179. That was confirmed for this one date. 

Now lets get a feel for the data, espeically in regard to plants. 

```{r, echo=FALSE, message=FALSE}

# Subset of all data that includes plants that were observed for flowering.
datphen2 %>% select(common_name, lifeform, event) %>% filter(lifeform == "PLANTS" & event =="FLOWERING") %>% group_by(common_name) %>% tally() %>% arrange(-n)

```

From the data, we see that there are 392 plants that were observed over the years. However, many have a sparse number of observations that are difficult to build models for. 

Here is a look at the data distribution



```{r, echo=FALSE, message=FALSE}

a<-datphen2 %>% select(common_name, lifeform, event) %>% filter(lifeform == "PLANTS" & event =="FLOWERING") %>% group_by(common_name) %>% tally() %>% arrange(-n)



  ggplot(a, mapping = aes(x = reorder(common_name, -n), y = n)) +
  geom_bar(colour="black", fill="blue", stat = "identity") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_text(size=16),
        axis.title=element_text(size=16,face="bold"),
        plot.title = element_text(size=22),
        axis.ticks.x=element_blank()) +
  labs(title = "Observations per plant species over 32 years", x = "Plant Species as Categories", y = "Number of observations per species")


```

We can see from the histogram that the majority of data are sparse indeed, with less than 10 observations. 


If we subset the data to those plants that have at least 15 or more observations, we still get 76 plant speices (see code below). this is a considerable number of plants for for trends in phenology. To interrogate the data further, lets restrict the phenology traits to flowering time only, and tidy the data up for recursive model fitting.



```{r, echo=FALSE, message=FALSE}

#subset for those speices that have 15 or more years of flowering observations
b<- a %>% filter(n >= 15)
c<-b %>% select(common_name)

#now make a dataframe that has all observation from only the species meeting out criteron (common name in dataframe b)

lm.data<-inner_join(c, datphen2, by = "common_name") %>% filter(event =="FLOWERING") #%>% group_by(common_name) %>% distinct(year, .keep_all = TRUE)

lm.data$year<-as.numeric(lm.data$year)


```

Now that the data are tidy (in a format to easily compute on), lets run a linear model across all 76 species through the 32 years of data. Further, lets extract the model slope and p.value to identify candidate species for further investigation on flowing time correlations with year. 

```{r, echo=FALSE, message=FALSE}

library(broom)

#running the lm model across all rows, genotypes
lm.results<- lm.data %>% group_by(common_name) %>% do(fitdata = lm(julian ~ year, data = .))

#getting tidy data output from model run
lmSlopePvalue <- tidy(lm.results, fitdata) %>% select(common_name, term, estimate, p.value) %>% filter(term =="year")

lmRsquare <- glance(lm.results, fitdata) %>% select(common_name, r.squared)

#tidy data output
lmtidyoutput<-left_join(lmSlopePvalue, lmRsquare, by = c("common_name" = "common_name"))

# lm model parameter distributions

par(mfrow=c(1,3))
hist(lmtidyoutput$estimate, main = "lm slope")
hist(lmtidyoutput$p.value, main = "P value")
hist(lmtidyoutput$r.squared, main = "r2")


```


The above histograms show distirbutions of linear model paramters, 


```{r, echo=FALSE, message=FALSE}
par(mfrow=c(1,3))
plot(lmtidyoutput$p.value, lmtidyoutput$estimate, main = "p.value to slope")
plot(lmtidyoutput$r.squared, lmtidyoutput$estimate, main = "r.squared to slope")
plot(lmtidyoutput$p.value, lmtidyoutput$r.squared, main = "p.value to r.squared")
```



















```{r, echo=FALSE, message=FALSE}

#looking at big toothed aspen flowering

datphen2$year<- as.numeric(datphen2$year)

datphen2 %>% filter(common_name=="BIG TOOTHED ASPEN" & event=="FLOWERING") %>% ggplot(aes(y= julian, x= year)) +
  geom_point()+
  geom_smooth(method = lm) +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=16,face="bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.text=element_text(size=16)) +
  labs(title = "Date of Big Toothed Aspen observed first flower", x = "Year", y = "Julian Calendar Day")

```


Not many observations between 1990 and 2000. To add confidence to the regression trend lets restrict the analysis to those years above 2000.


```{r, echo=FALSE, message=FALSE}

datphen2 %>% filter(year > 2000 & common_name=="BIG TOOTHED ASPEN" & event=="FLOWERING") %>% ggplot(aes(y= julian, x= year)) +
  geom_point()+
  geom_smooth(method = lm) +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=16,face="bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.text=element_text(size=16)) +
  labs(title = "Date of Big Toothed Aspen fist flower", x = "Year", y = "Julian Calendar Day")


```


The trend between the two above graphs is similar, with trends showing a decrase in flowering time by year. 

now lets look at trends across other plants, do they all show the same trend?

We arbitrarily pick 15 observations at the minimum number.

```{r, echo=FALSE, message=FALSE}


#subset for those speices that have more than 15 years of flwoering observations
b<- a %>% filter(n > 15)
c<-b %>% select(common_name)

#now make a dataframe that has all observation from only the species meeting out criteron (common name in dataframe b)

lm.data<-inner_join(c, datphen2, by = "common_name") %>% filter(event =="FLOWERING")

```

We can see that dataframe b has 67 rows, meaning that only 67 plant species have 15 or more observations for this analysis. 


Now lets write code for a model that looks for regression trends among flowering and year for all plant species meeting a 15 year minimum criteron. 


```{r, echo=FALSE, message=FALSE}

library(broom)

#running the lm model across all rows, genotypes
lm.results<- lm.data %>% group_by(common_name) %>% do(fitdata = lm(julian ~ year, data = .))

#getting tidy data output from model run
lmSlopePvalue <- tidy(lm.results, fitdata) %>% select(common_name, term, estimate, p.value) %>% filter(term =="year")

lmRsquare <- glance(lm.results, fitdata) %>% select(common_name, r.squared)

#tidy data output
lmtidyoutput<-left_join(lmSlopePvalue, lmRsquare, by = c("common_name" = "common_name"))

# lm model parameter distributions

par(mfrow=c(1,3))
hist(lmtidyoutput$estimate, main = "lm slope")
hist(lmtidyoutput$p.value, main = "P value")
hist(lmtidyoutput$r.squared, main = "r2")


```



```{r, echo=FALSE, message=FALSE}
par(mfrow=c(1,3))
plot(lmtidyoutput$p.value, lmtidyoutput$estimate, main = "p.value to slope")
plot(lmtidyoutput$r.squared, lmtidyoutput$estimate, main = "r.squared to slope")
plot(lmtidyoutput$p.value, lmtidyoutput$r.squared, main = "p.value to r.squared")
```


```{r, echo=FALSE, message=FALSE}

#now lets graph the data and see what it looks like
#select the p.value cutoff

trash<-lmtidyoutput %>% filter(p.value <=0.05) %>% select(common_name)

trash2<-left_join(trash, datphen2, by = c("common_name" = "common_name"))

ggplot(trash2, aes(y= julian, x= year, colour = common_name, group = common_name)) +
  geom_point()+
  geom_quantile() +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=16,face="bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.text=element_text(size=16)) 

#determine how many species meet our criteria


     

trash3<-trash2 %>% filter(event == "FLOWERING")
    

## start of plotting function

plant.species <- function(df, na.rm = TRUE, ...){
  
  # create list of counties in data to loop over 
  plant_list <- unique(df$common_name)
  
  # create for loop to produce ggplot2 graphs 
  for (i in seq_along(plant_list)) { 
    
    # create plot for each county in df 
    plot <- 
      ggplot(subset(df, df$common_name==plant_list[i]),
        aes(y= julian, x= year, group = common_name)) + 
        geom_point(size=4, colour = "dark blue")+
        geom_smooth(method = lm) +
        theme(axis.text=element_text(size=12),
        axis.title=element_text(size=16,face="bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.text=element_text(size=16)) +
        ggtitle(paste(plant_list[i]))
    
    # save plots as .png
    # ggsave(plot, file=paste(results, 
    #                        'projection_graphs/county_graphs/',
    #                        plant_list[i], ".png", sep=''), scale=2)
   
    # save plots as .pdf
    # ggsave(plot, file=paste(results, 
    #                        'projection_graphs/county_graphs/',
    #                        plant_list[i], ".pdf", sep=''), scale=2)
    
    # print plots to screen
    print(plot)
  }
}

# run graphing function on long df
plant.species(trash3)

```


A closer look at False Solomon Seal

```{r, echo=FALSE, message=FALSE}

datphen2 %>% filter(common_name=="FALSE SOLOMON SEAL" & event=="FLOWERING") %>% ggplot(aes(y= julian, x= year)) +
  geom_point(size=4, colour = "dark blue")+
  geom_smooth(method = lm) +
  theme(axis.text=element_text(size=18),
        axis.title=element_text(size=20,face="bold"),
        plot.title = element_text(size=20, face = "bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.text=element_text(size=20)) +
  labs(title = "Date of False Somolon's Seal observed first flower", x = "Year", y = "Julian Calendar Day")


```



Using the code above you can explore the Latimer bird data in a similar way. If you do so, you will notice the bird data is much more sparse than than the plant. Suggesting that Latimer was more interested in observing plant phenology than bird phenology. Below is an example graph chosen because of numerous observations (abundant data) and seeminly non-responive trend. An informed public must keep in mind that graphs showing showing exagerated trends may not be the norm, indeed the DarK Eyed Junco seems to be non-responsive to year observed - see below


```{r, echo=FALSE, message=FALSE}
datphen2 %>% filter(common_name=="DARK EYED JUNCO" & event=="FIRST SEEN") %>% ggplot(aes(y= julian, x= year)) +
  geom_point(size=4, colour = "dark blue")+
  geom_smooth(method = lm) +
  theme(axis.text=element_text(size=18),
        axis.title=element_text(size=20,face="bold"),
        plot.title = element_text(size=20, face = "bold"),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.text=element_text(size=20)) +
        scale_y_continuous(limits = c(0, 200)) +
  labs(title = "Date when Dark Eyed Junco was first seen", x = "Year", y = "Julian Calendar Day")
```






